---
title: "Load and explore the data"
teaching: 15
exercises: 30
questions:
- "What data are required for eqtl mapping?"
objectives:
- To provide an example and exploration of data used for eqtl mapping.
keypoints:
- ""
source: Rmd
---

```{r, include=FALSE}
source("../bin/chunk-options.R")
knitr_fig_path("03-")
```

Load the libraries.
```{r, load_tidyverse, message=FALSE}
library(tidyverse)
library(knitr)
library(corrplot)
# the following analysis is derived from supplementary 
# File S1 Attie_eQTL_paper_physiology.Rmd 
# by Daniel Gatti. See Data Dryad entry for more information.
```

## Physiological Phenotypes

The complete data used in these analyses are available from 
[Data Dryad](https://doi.org/10.5061/dryad.pj105). 

Load in the clinical phenotypes.

```{r load_phenos}
# load the data
load("../data/attie_DO500_clinical.phenotypes.RData")
```

See the [data dictionary](../data/Attie-232_Attie_DO_Islets-dictionary.csv) to 
see a description of each of these phenotypes. You can also view a table of
the data dictionary.

```{r dict_phenos}
pheno_clin_dict %>% 
  select(description, formula) %>% 
  kable()
```

### Boxplots
Boxplots are a great way to view the distribution of the data and to identify 
any outliers. The following boxplots display repeated measurements of body 
weight, glucose, insulin and triglyceride.

```{r bw_boxplot, warning=FALSE}
# pull out body weights over time and place in long form
body_weights <- pheno_clin %>%
  select(mouse, sex, starts_with("weight")) %>%
  pivot_longer(cols=starts_with("weight"), 
               names_to = "week",
               values_to = "weight", 
               names_transform = list(week = readr::parse_number))

# redefine week as a factor so that plot will render correctly
body_weights$week <- as.factor(body_weights$week)

# plot body weights on a log 10 scale
ggplot(body_weights, aes(week, weight, fill = sex)) +
  geom_boxplot() +
  scale_y_log10() +
  labs(title = "Body Weight", y = "Weight (g)")
```

Repeat boxplots for glucose, insulin and triglyceride measurements in the same 
manner as for weight.

```{r glucose_boxplot, warning=FALSE}
glucose <- pheno_clin %>%
  select(mouse, sex, (starts_with("Glu_") & ends_with("wk"))) %>%
  pivot_longer(cols=starts_with("Glu"), 
               names_to = "week",
               values_to = "glucose", 
               names_transform = list(week = readr::parse_number))

# redefine week as a factor so that plot will render correctly
glucose$week <- as.factor(glucose$week)

ggplot(glucose, aes(week, glucose, fill = sex)) +
  geom_boxplot() +
  scale_y_log10() +
  labs(title = "Fasting Plasma Glucose", y = "Glucose (mg/dl)")

```

```{r insulin_boxplot, warning=FALSE}
insulin <- pheno_clin %>%
  select(mouse, sex, (starts_with("Ins_") & ends_with("wk"))) %>%
  pivot_longer(cols=starts_with("Ins"), 
               names_to = "week",
               values_to = "insulin", 
               names_transform = list(week = readr::parse_number))
# redefine week as a factor so that plot will render correctly
insulin$week <- as.factor(insulin$week)

ggplot(insulin, aes(week, insulin, fill = sex)) +
  geom_boxplot() +
  scale_y_log10() +
  labs(title = "Fasting Plasma Insulin", y = "Insulin (ng/ml)")
```

```{r trig_boxplot, warning=FALSE}
triglyceride <- pheno_clin %>%
  select(mouse, sex, starts_with("TG")) %>%
  pivot_longer(cols=starts_with("TG"), 
               names_to = "week",
               values_to = "triglyceride", 
               names_transform = list(week = readr::parse_number))

# redefine week as a factor so that plot will render correctly
triglyceride$week <- as.factor(triglyceride$week)

ggplot(triglyceride, aes(week, triglyceride, fill = sex)) +
  geom_boxplot() +
  scale_y_log10() +
  labs(title = "Fasting Plasma Triglyceride", y = "Triglyceride (mg/dl)")
```

```{r fig1_boxplots, warning=FALSE}
remaining_phenos <- pheno_clin %>%
  select(mouse, sex, diet_days:Ins_tAUC, food_ave) %>% 
  pivot_longer(!c(mouse, sex), names_to = "phenotype", values_to = "value")
  
ggplot(remaining_phenos, aes(sex, value, fill = sex)) +
    geom_boxplot() +
    scale_y_log10() +
    facet_wrap(~phenotype, scales = "free_y")
```

### QA/QC

Log transform and standardize each phenotype. Consider setting points that are 
more than 5 std. dev. from the mean to NA. Only do this if the final 
distribution doesn't look skewed.

```{r pheno_std, warning=FALSE}
pheno_clin_log = pheno_clin %>%
                   mutate_if(is.numeric, log)
pheno_clin_std = pheno_clin_log %>%
                   select(mouse, num_islets:weight_10wk) %>%
                   mutate_if(is.numeric, scale)
pheno_clin_std %>%
  select(num_islets:weight_10wk) %>%
  gather(phenotype, value) %>%
  ggplot(aes(x = phenotype, y = value)) +
    geom_boxplot() +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
    labs(title = "Distribution of Standardized Phenotypes")
```

### Tests for sex, wave and diet_days.

```{r convert_factors}
# convert sex and DO wave (batch) to factors
pheno_clin$sex = factor(pheno_clin$sex)
pheno_clin$DOwave = factor(pheno_clin$DOwave)
```

```{r sex_diet_wave_anova}
tmp = pheno_clin_log %>%
        select(mouse, sex, DOwave, diet_days, num_islets:weight_10wk) %>%
        gather(phenotype, value, -mouse, -sex, -DOwave, -diet_days) %>%
        group_by(phenotype) %>%
        nest()
mod_fxn = function(df) {
  lm(value ~ sex + DOwave + diet_days, data = df)
}
tmp = tmp %>%
  mutate(model = map(data, mod_fxn)) %>%
  mutate(summ = map(model, tidy)) %>%
  unnest(summ)
# kable(tmp, caption = "Effects of Sex, Wave & Diet Days on Phenotypes")
```

```{r sex_diet_wave_effects}
tmp %>%
  filter(term != "(Intercept)") %>%
  mutate(neg.log.p = -log10(p.value)) %>%
  ggplot(aes(term, neg.log.p)) +
    geom_point() +
    facet_wrap(~phenotype) +
    labs(title = "Significance of Sex, Wave & Diet Days on Phenotypes") +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) +
rm(tmp)
```

### Weight vs. Food Intake

```{r bw_vs_food, warning=FALSE, message=FALSE}
pheno_clin_log %>%
  select(mouse, sex, food_ave:weight_10wk) %>%
  gather(phenotype, value, -mouse, -sex, -food_ave) %>%
  separate(phenotype, c("phenotype", "week")) %>%
  mutate(week = factor(week, levels = c("2wk", "6wk", "10wk"))) %>%
  ggplot(aes(food_ave, value, color = sex)) +
    geom_point() +
    geom_smooth(method = "lm") +
    labs(title = "Food Intake vs. Body Weight", y = "log(Body Weight)") + 
    facet_wrap(~week)
```

```{r bw_sex_food_model}
model_fxn = function(df) { lm(value ~ sex*food_ave, data = df) }
tmp = pheno_clin_log %>%
  select(mouse, sex, food_ave:weight_10wk) %>%
  gather(phenotype, value, -mouse, -sex, -food_ave) %>%
  separate(phenotype, c("phenotype", "week")) %>%
  mutate(week = factor(week, levels = c("2wk", "6wk", "10wk"))) %>%
  group_by(week) %>%
  nest() %>%
  mutate(model = map(data, model_fxn)) %>%
  mutate(summ = map(model, tidy)) %>%
  unnest(summ) %>%
  filter(term != "(Intercept)") %>%
  mutate(p.adj = p.adjust(p.value))
# kable(tmp, caption = "Effects of Sex and Food Intake on Body Weight")
```

### Correlation Plots

Females

```{r female_corr_plot, fig.width=15, fig.height=15, message=FALSE, warning=FALSE}
tmp = pheno_clin_log %>% 
        filter(sex == "F") %>%
        select(starts_with(c("Ins", "Glu", "TG")))
tmp = cor(tmp, use = "pairwise")
corrplot.mixed(tmp, upper = "ellipse", lower = "number", 
               main = "Female Clinical Phenotype Correlation")
corrplot.mixed(tmp, upper = "ellipse", lower = "number", 
               main = "Female Clinical Phenotype Correlation")
```

Males

```{r male_corr_plot, fig.width=15, fig.height=15, warning=FALSE, message=FALSE}
tmp = pheno_clin_log %>% 
        filter(sex == "M") %>%
        select(starts_with(c("Ins", "Glu", "TG")))
tmp = cor(tmp, use = "pairwise")
corrplot.mixed(tmp, upper = "ellipse", lower = "number", 
               main = "Male Clinical Phenotype Correlation")
corrplot.mixed(tmp, upper = "ellipse", lower = "number", 
               main = "Male Clinical Phenotype Correlation")
```

## Gene Expression Phenotypes
```{r load_expression_data}
# load the expression data along with annotations and metadata
load("../data/dataset.islet.rnaseq.RData")
names(dataset.islet.rnaseq)
```

```{r view_annot_rawCounts_metadata}
# look at gene annotations
dataset.islet.rnaseq$annots[1:6,]

# look at raw counts
dataset.islet.rnaseq$raw[1:6,1:6]

# look at sample metadata
# summarize mouse sex, birth dates and DO waves
table(dataset.islet.rnaseq$samples[, c("sex", "birthdate")])
table(dataset.islet.rnaseq$samples[, c("sex", "DOwave")])
```

In order to make reasonable gene comparisons between samples, the count data 
need to be normalized. In the quantile-quantile (Q-Q) plot below, count data for 
the first gene are plotted over a diagonal line tracing a normal distribution 
for those counts. Notice that most of the count data values lie off of this 
line, indicating that these gene counts are not normally distributed. 

```{r view_manual_qqplot_raw, echo=FALSE}
library(rafalib)
ps <- ( seq(0,99) + 0.5 )/100
qs <- quantile(dataset.islet.rnaseq$raw[,1], ps)
normalqs <- qnorm(ps, 
                  mean(dataset.islet.rnaseq$raw[,1]),
                  popsd(dataset.islet.rnaseq$raw[,1]))
plot(normalqs, qs, 
     xlab="Normal percentiles", ylab="Count percentiles",
     main="Count distribution for gene ENSMUSG00000000001")
abline(0,1) # identity line
```

The same plot can be drawn as shown below. The diagonal line represents a 
standard normal distribution with mean 0 and standard deviation 1. Count data
values are plotted against this standard normal distribution.

```{r view_qqplot_raw}
qqnorm(dataset.islet.rnaseq$raw[,1], 
     xlab="Theoretical normal percentiles", 
     ylab="Count percentiles",
     main="Count distribution for gene ENSMUSG00000000001")
qqline(dataset.islet.rnaseq$raw[,1]) 
```

Q-Q plots for the first six genes show that count data for these genes are not
normally distributed. They are also not on the same scale. The y-axis values for
each subplot range to 20,000 counts in the first subplot, 250 in the second, 80
in the third, and so on. 

```{r view_qqplots_raw}
par(mfrow=c(2,3))
for (i in 1:6) {
  qqnorm(dataset.islet.rnaseq$raw[,i])
  qqline(dataset.islet.rnaseq$raw[,i])
  }
```

Q-Q plots of the normalized expression data for the first six genes show that 
the data values match the diagonal line well, meaning that they are now normally
distributed. They are also all on the same scale now as well.

```{r view_qqplots_normalized}
par(mfrow=c(2,3))
for (i in 1:6) {
  qqnorm(dataset.islet.rnaseq$expr[,i])
  qqline(dataset.islet.rnaseq$expr[,i])
  }
```

Boxplots of raw counts for 5 example genes are shown at left below. Notice that 
the median count values (horizontal black bar in each boxplot) are not 
comparable between the genes because the counts are not on the same scale. At
right, boxplots for the same 5 genes show normalized count data on the same 
scale.

```{r view_example_boxplots}
par(las=2, mfrow=c(1,2))
boxplot(dataset.islet.rnaseq$raw[,c(5:9)], 
        main="Raw count distributions for 5 example genes")
boxplot(dataset.islet.rnaseq$expr[,c(5:9)], 
        main="Normalized count distributions for the same 5 example genes")
```

Have a look at the first several rows of normalized count data.

```{r normalized_counts}
# look at normalized counts
dataset.islet.rnaseq$expr[1:6,1:6]
```

The expression data loaded provides LOD peaks for the eQTL analyses performed in
this study. As a preview of what you will be doing next, look at the first 
several rows of LOD peak values and extract the LOD peaks for chromosome 11.

```{r lod_peaks}
# look at LOD peaks
dataset.islet.rnaseq$lod.peaks[1:6,]

# look at chromosome 11 LOD peaks
chr11_peaks <- dataset.islet.rnaseq$annots %>% 
   select(gene_id, chr) %>% 
   filter(chr=="11") %>%
   left_join(dataset.islet.rnaseq$lod.peaks, 
             by = c("chr" = "chrom", "gene_id" = "annot.id")) 

# look at the first several rows of chromosome 11 peaks
head(chr11_peaks)

# how many rows?
dim(chr11_peaks)

# how many rows have LOD scores?
chr11_peaks %>% filter(!is.na(lod)) %>% dim()

# sort chromosome 11 peaks by LOD score
chr11_peaks %>% arrange(desc(lod)) %>% head()

# range of LOD scores and positions
range(chr11_peaks$lod, na.rm = TRUE)
range(chr11_peaks$pos, na.rm = TRUE)

# view LOD scores by position
chr11_peaks %>% arrange(desc(lod)) %>% 
  ggplot(aes(pos, lod)) + geom_point()
```